Safe Redis Queries

### 2016-06-11 aplas16/main-aplas.tex
Revision up to Section 2.

\usepackage{amsmath}
%\usepackage{amsfonts}
%\usepackage{amsthm}
\usepackage{url}
\usepackage{mathptmx}
\usepackage{doubleequals}
\usepackage{makeidx}  % allows for indexgeneration

@@

\title{Preventing Runtime Errors of Redis at Compile Time}
\author{Ting-Yan Lai \inst{1}\inst{2}\and Tyng-Ruey Chuang \inst{1}
  \and Shin-Cheng Mu \inst{1}}
\institute{Institute of Information Science, Academia Sinica, Taiwan
\and Institute of Computer Science and Engineering,
National Chiao Tung University, Taiwan}

@@

\begin{abstract}
Programmers often interact with database systems by sending queries through libraries or packages in some programming languages. People, however, make mistakes.
While some of the syntactic and semantic errors can be prevented by the
 language at compile time, or caught by the package at runtime, most semantic
 errors are just being ignored, causing problems at the database system.

In this paper, we demonstrate how to prevent those runtime errors at compile
 time, thus allowing users to write more reliable database queries without
 runtime overhead, by exploiting type-level programming techniques such as
 indexed monad, type-level literals and closed type families in Haskell.

The database system and the package we are targeting are \emph{Redis} and
 \emph{Hedis} respectively, and our implementation is available as \emph{Popcorn}
 on \emph{Hackage}.
\end{abstract}

\input{sections/Introduction}
\input{sections/Motivation}
\input{sections/TypeLevelDict}
\input{sections/IndexedMonads}
\input{sections/TypeLevelFun}
\input{sections/PolyRedis}
\input{sections/Constraints}
\input{sections/Assertions}
\input{sections/Discussions}
\input{sections/Related}
\input{sections/Conclusions}

>>>>

\usepackage{amsmath}
%\usepackage{amsfonts}
%\usepackage{amsthm}
\usepackage{color}
\usepackage{url}
\usepackage{mathptmx}
\usepackage{doubleequals}
\usepackage{makeidx}  % allows for indexgeneration

@@

\title{Safe Redis Queries}
\subtitle{A Case Study of Type-Level Programming}
\author{Ting-Yan Lai \inst{1}\inst{2}\and Tyng-Ruey Chuang \inst{1}
  \and Shin-Cheng Mu \inst{1}}
\institute{Institute of Information Science, Academia Sinica, Taiwan
\and Institute of Computer Science and Engineering,
National Chiao Tung University, Taiwan}

@@

\begin{abstract}
\Redis{} is an in-memory data strcture store, often used as a database, with a
Haskell interface \Hedis{}. \Redis{} is dynamically typed --- a value associated
with a key can be overwritten by a value of a different type, and a command,
when fetching a value of a type it does not expect, signals a runtime error. We
develop a domain-specific language that, by exploiting Haskell type-level
programming techniques including indexed monad, type-level literals and closed
type families, keeps track of types of values in the database and statically
guarantees that type errors cannot happen.
\end{abstract}

\input{sections/Introduction}
%\input{sections/Motivation}
\input{sections/TypeLevelDict}
\input{sections/IndexedMonads}
\input{sections/TypeLevelFun}
\input{sections/PolyRedis}
\input{sections/Constraints}
\input{sections/Assertions}
\input{sections/Discussions}
\input{sections/Related}
\input{sections/Conclusions}

### 2016-06-11 aplas16/sections/Introduction.tex
Revision up to Section 2.

\subsection{Redis}

Redis\footnote{\url{https://redis.io}} is an open source, in-memory data structure store, often used as
database, cache and message broker. A Redis datatype can be think of as a set of key-value pairs, where each value is associated with a binary-safe string key to identify and manipulate with. Redis supports many different kind of values, such as strings, hashes, lists, and sets, etc, and provides a collection of of atomic \emph{commands} to manipulate these values.

For an example, consider the following sequence of commands, entered through the interactive interface of Redis. The keys
\text{some-set} abd \text{another-set} are
both associated to a set. The two call to command \text{SADD} respectively adds three and two values to the two sets, before \text{SINTER} takes their intersection:
\begin{tabbing}\tt
~redis\char62{}~SADD~some\char45{}set~a~b~c\\
\tt ~\char40{}integer\char41{}~3\\
\tt ~redis\char62{}~SADD~another\char45{}set~a~b\\
\tt ~\char40{}integer\char41{}~2\\
\tt ~redis\char62{}~SINTER~some\char45{}set~another\char45{}set\\
\tt ~1\char41{}~\char34{}a\char34{}\\
\tt ~2\char41{}~\char34{}b\char34{}
\end{tabbing}

Note that the keys \text{some-set} and \text{another-set}, if not existing before the call to \text{SADD}, are created on site. The call to
\text{SADD} returns the size of the set after
completion of the command.

\subsection{Hedis}

Many third party library exist to allow general purpose programmings
to access Redis databases through its TCP protocol. The most popular
such library for Haskell is Hedis\footnote{\url{https://hackage.haskell.org/package/hedis}}.

The following program is how the previous example looks like in
Hedis:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{program}\mathbin{::}\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;[\mskip1.5mu \Conid{ByteString}\mskip1.5mu]){}\<[E]%
\\
\>[B]{}\Varid{program}\mathrel{=}\mathbf{do}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{sadd}\;\text{\tt \char34 some-set\char34}\;[\mskip1.5mu \text{\tt \char34 a\char34},\text{\tt \char34 b\char34}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{sadd}\;\text{\tt \char34 another-set\char34}\;[\mskip1.5mu \text{\tt \char34 a\char34},\text{\tt \char34 b\char34},\text{\tt \char34 c\char34}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{sinter}\;[\mskip1.5mu \text{\tt \char34 some-set\char34},\text{\tt \char34 another-set\char34}\mskip1.5mu]{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The function \ensuremath{\Varid{sadd}} takes a key and a list of values as
 arguments, and returns an \ensuremath{\Conid{Integer}} on success, or
 returns a \ensuremath{\Conid{Reply}}, a low-level representation of replies
 from the Redis server, in case of failures. All wrapped in
 \ensuremath{\Conid{Redis}}\footnotemark, the context of command execution.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{27}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{sadd}{}\<[7]%
\>[7]{}\mathbin{::}\Conid{ByteString}{}\<[27]%
\>[27]{}\mbox{\onelinecomment  key}{}\<[E]%
\\
\>[7]{}\to [\mskip1.5mu \Conid{ByteString}\mskip1.5mu]{}\<[27]%
\>[27]{}\mbox{\onelinecomment  values}{}\<[E]%
\\
\>[7]{}\to \Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

\footnotetext{Hedis provides another kind of context, \ensuremath{\Conid{RedisTx}}, for \emph{transactions}, united with \ensuremath{\Conid{Redis}} under the class of \ensuremath{\Conid{RedisCtx}}. For brevity, we demonstrate only \ensuremath{\Conid{Redis}} in this paper. }

>>>>

\Redis{}\footnote{\url{https://redis.io}} is an open source, in-memory data structure store, often used as database, cache and message broker. A \Redis{} datatype can be think of as a set of key-value pairs, where each value is associated with a binary-safe string key to identify and manipulate with.
\Redis{} allows values of various types, including strings, hashes, lists, and sets, etc, to be stored, and provides a collection of of atomic \emph{commands} to manipulate these values.

For an example, consider the following sequence of commands, entered through the interactive interface of \Redis{}. The keys \texttt{some-set} and \texttt{another-set}
are both associated to a set. The two call to command \texttt{SADD} respectively
adds three and two values to the two sets, before \texttt{SINTER} takes their intersection:
\begin{tabbing}\tt
~redis\char62{}~SADD~some\char45{}set~a~b~c\\
\tt ~\char40{}integer\char41{}~3\\
\tt ~redis\char62{}~SADD~another\char45{}set~a~b\\
\tt ~\char40{}integer\char41{}~2\\
\tt ~redis\char62{}~SINTER~some\char45{}set~another\char45{}set\\
\tt ~1\char41{}~\char34{}a\char34{}\\
\tt ~2\char41{}~\char34{}b\char34{}
\end{tabbing}

Note that the keys \texttt{some-set} and \texttt{another-set}, if not existing before the call to \texttt{SADD}, are created on site. The call to
\texttt{SADD} returns the size of the set after completion of the command.

Many third party libraries provide interfaces that allow general purpose programming languages to access \Redis{} through its TCP protocol.
For Haskell, the most popular library is \Hedis{}\footnote{\url{https://hackage.haskell.org/package/hedis}}.
The following program implements the previous example:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{program}\mathbin{::}\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;[\mskip1.5mu \Conid{ByteString}\mskip1.5mu]){}\<[E]%
\\
\>[B]{}\Varid{program}\mathrel{=}\mathbf{do}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{sadd}\;\text{\tt \char34 some-set\char34}\;[\mskip1.5mu \text{\tt \char34 a\char34},\text{\tt \char34 b\char34}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{sadd}\;\text{\tt \char34 another-set\char34}\;[\mskip1.5mu \text{\tt \char34 a\char34},\text{\tt \char34 b\char34},\text{\tt \char34 c\char34}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{sinter}\;[\mskip1.5mu \text{\tt \char34 some-set\char34},\text{\tt \char34 another-set\char34}\mskip1.5mu]~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The function \ensuremath{\Varid{sadd}\mathbin{::}\Conid{ByteString}\to [\mskip1.5mu \Conid{ByteString}\mskip1.5mu]\to \Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer})} takes a key and a list of values as arguments, and returns
an \ensuremath{\Conid{Integer}} on success, or returns a \ensuremath{\Conid{Reply}}, a low-level representation of
replies from the Redis server, in case of failures. All wrapped in the monad
\ensuremath{\Conid{Redis}}, the context of command execution.\footnotemark

@@

\footnotetext{\Hedis{} provides another kind of context, \ensuremath{\Conid{RedisTx}}, for \emph{transactions}, united with \ensuremath{\Conid{Redis}} under the class \ensuremath{\Conid{RedisCtx}}. We
demonstrate only \ensuremath{\Conid{Redis}} in this paper.}

%\paragraph{Motivation}
%All binary strings are equal, but some binary strings are more equal than others.
%While everything in \Redis{} is essentially a binary string, these strings
%are treated differently.
\Redis{} supports many different kind of data
structures, such as strings, hashes, lists, etc. While they are all encoded as
binary strings before being written to the database, most commands only works
with data of certain types. In the following example, the key
\texttt{some-string} is associated to string \texttt{foo} --- the command
\texttt{SET} always associates a key to a string. The subsequent call to \texttt{SADD}, which adds a value to a set, thus causes a runtime error.
\begin{tabbing}\tt
~redis\char62{}~SET~some\char45{}string~foo\\
\tt ~OK\\
\tt ~redis\char62{}~SADD~some\char45{}string~bar\\
\tt ~\char40{}error\char41{}~WRONGTYPE~Operation~against~a~key~holding~the~wrong\\
\tt ~kind~of~value
\end{tabbing}

%\paragraph{Example 2} Even worse, not all strings are equal!
% The call \texttt{INCR some-string} parses the string associated with key
% \texttt{some-string} to an integer, increments it by one, and store it back as
% a string. If the string can not be parse as an integer, a runtime error
% is raised.
% \begin{verbatim}
% redis> SET some-string foo
% OK
% redis> INCR some-string
% (error) ERR value is not an integer or out of range
% \end{verbatim}

Being a simple wrapper on top of the TCP protocol of \Redis{}, \Hedis{}
inherits the problem. Executing following program yields the same error
wrapped in Haskell: \ensuremath{\Conid{Left}\;(\Conid{Error}} \texttt{"WRONGTYPE Operation against a
key holding the wrong kind of value"}\ensuremath{)}.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{program}\mathbin{::}\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
\\
\>[B]{}\Varid{program}\mathrel{=}\mathbf{do}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{set}\;\text{\tt \char34 some-string\char34}\;\text{\tt \char34 foo\char34}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{sadd}\;\text{\tt \char34 some-string\char34}\;[\mskip1.5mu \text{\tt \char34 a\char34}\mskip1.5mu]~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

% \paragraph{The Cause} Every key is associated with a value, and every value has
% its own type. But most commands in \Redis{} only work with a certain type of
%  value. When a command is used on a wrong type of key, a runtime error occurs.
%  The problems illustrated above arise from the absence of type checking, with
%  respects to \textbf{the type of a value that associates with a key}.
%  These problems could have been avoided, if we could know the type every key
%  associates with in advance, and prevent programs with invalid commands from
%  executing.
%
% \paragraph{Hedis as an embedded DSL}
% Haskell makes it easy to build and use {\em domain specific embedded languages} (DSELs), and \Hedis{} can be regarded as one of them. What makes \Hedis{} peculiar is that,
%  it has \emph{variable bindings} (between keys and values), but with very
%  little or no semantic checking, neither dynamically nor statically.
%
Such a programming model is certainly very error-prone. Working within Haskell,
a host language with a strong typing system, one naturally wishes to build a
a domain-specific embedded language (DSEL) that exploits the rich type system
of Haskell to ensure absence of \Redis{} type errors.

This paper discusses the techniques we used and experiences we learned from building such a language, nicknamed \Popcorn{}. We constructed an {\em indexed
monad}, on top of the monad \ensuremath{\Conid{Redis}}, which is indexed by a dictionary that
maintains the set of currently defined keys and their types. To represent
the dictionary, we need to encode variable binds with {\em type-level} lists
and strings. To summarize our contributions:
\begin{itemize}
\item We present \Popcorn{}, a statically typed domain-specific language embedded in Haskell and built on \Hedis{}.
% also makes Redis polymorphic by automatically converting back and forth from values of arbitrary types and boring ByteStrings.
%
\item We demonstrate how to model variable bindings of an embedded DSL using
 language extensions including type-level literals and data kinds.
%
\item We provide (yet another) an example of encoding effects and constraints of
in types, with indexed monad~\cite{indexedmonad}, closed type-families~\cite{closedtypefamilies} and constraints kinds~\cite{constraintskinds}.
\end{itemize}
\todo{Phrase this better.}

### 2016-06-11 aplas16/sections/TypeLevelDict.tex
Revision up to Section 2.

To check the bindings between keys and values, we need a \emph{dictionary-like}
 structure, and encode it as a \emph{type} somehow.

\subsection{Datatype promotion}
Normally, at the term level, we could express the datatype of dictionary with
\emph{type synonym} like this.\footnotemark

>>>>

One of the challenges of statically ensuring type correctness of \Redis{},
which also presents in other stateful languages, is that the type of the value
associated to a key can be altered after updating. To ensure type correctness,
we have to keep track of the (\Redis{}) types of all existing keys in a
{\em dictionary} --- conceptually, a list of pairs of keys and \Redis{} types.
Each \Redis{} command is embedded in \Popcorn{} as a monadic computation. The
monad, to be presented in Section \ref{sec:indexed-monads}, is indexed by
the dictionaries before and after the computation. In a dependently typed
programming language (without the so-called ``phase distinction'' ---
separation between types and terms), this would pose no problem. In Haskell
however, the dictionaries, to index a monad, has to be a Haskell type as well.

In this section we describe how to construct a type-level dictionary, to be
used with the indexed monad in Section \ref{sec:indexed-monads}. More operations
on the dictionary will be presented in Section \ref{sec:type-level-fun}.

\subsection{Datatype Promotion}

Normally, at the term level, we could express the datatype of dictionary with
\emph{type synonym} like this.\footnotemark
